Causal Impact of Masks, Policies, Behavior on Early Covid-19 Pandemic in the U.S.

### 2020-09-04 tex/reply-to-referees.tex

\newcommand{\replytitle}
{
\subsection*{Reply to the Comments on  ``Causal Impact of Masks, Policies, Behavior on Early COVID-19 Pandemic in The U.S.''}
}

\newcommand{\refreplyheading}
{Thank you very much for your helpful and constructive comments on our paper. We appreciate the time you spent in reviewing this paper. We have followed your recommendations, and those of the other reviewers', as far as possible.
Major changes are:}

\newcommand{\changes}
{
\begin{enumerate}

\item   We  state the limitations of our work with cautious notes on interpreting our causal model.
\item   Following the editor's suggestion, Section 2 is reorganized and we improve our discussion on the testable restrictions. Tables 7 and 9 in Section 3 provide   an  over-identification test for no indirect effect of mask policy on case/death growth via behavior.
\item   Following suggestions by the editor and the referees, we provide various sensitivity checks as well as the results of debiased fixed effects estimator  in Section 4.

\item Given Referee 1's comment, all regressions now include a dummy for state governor's party and its interaction with month dummies as additional controls.

\item   We improve the figures for counterfactual experiments so that they are now easier to understand.

\item We now combine movie theater closures, restaurant closures, and non-essential business closures as one policy variable called ``business closure policies'' by taking the average of these three variables. Accordingly, we conduct a counterfactual of keeping all business open.

\item We changed the starting date of mask mandates to March 14th in our counterfactual experiment of mandatory mask policy.

\item  We adjusted the timing of monthly dummies for estimating behavior equation ($PI \to B$)  in Table 3.  In the previous version, the definition of monthly dummies for behavior equation ($PI \to B$) is different from that for case growth equation ($BPI \to Y$)  or ($PI \to Y$). This issue is now fixed.
%\item We made some grammatical adjustments.
\end{enumerate}

@@

 We also note that the estimated coefficients of  $\Delta \log\Delta D_{it} $ and $\log\Delta D_{it}$ are negative in columns (1)-(4) of  Table 3(B), consistent with a hypothesis that a higher number of deaths as well as a higher growth rate of deaths reduces people's mobility.

Following your suggestion, we added the above discussion in footnote X (check!) when we discuss in Table 3A.

\item[11.]  \textit{Figure 8 is very informative. Perhaps the authors could show this in introduction as well.''}

We thought about this suggestion but, eventually, we can keep it as it is.

@@

We believe that it is better to use the number of cases and deaths rather than time since start of the outbreak as a  general framework to analyze the spread of COVID-19. This is especially true if we analyze the second wave of COVID-19, where time since start of the first outbreak may not be so relevant, and identifying the start of the second outbreak may be difficult.

\item \textit{`` Figure 8 - I do like seeing the data, but it also just raises questions, as you say, confounders + many other things at play.... I wondered if there would be a way to create a visual that captured a bit more of this, and pinpointed the anomalies which are where the inferential power lies. ''}

This is an interesting suggestion. We thought about the possibility of providing a version of Figure 8 based on the regression residuals after taking out the effect of confounders etc but, eventually, we decided not to present such results due to the space constraint.

The following figures are generated by first regressing the case or death growth rates on state characteristics, month dummies, and their interactions and then computing their average value for states with or without a mask mandate for each date.

\begin{figure}[ht]
  \caption{Average  residuals of regressing case and death growth on confounders among states with or without  mask mandates \label{fig:masks-res}}\vspace{0.2cm}
 \begin{minipage}{\linewidth}
    \centering
    \begin{tabular}{cc}
      \includegraphics[width=0.5\textwidth]{tables_and_figures/pmaskbus-cases-res-14}
      &
        \includegraphics[width=0.5\textwidth]{tables_and_figures/pmaskbus-deaths-res-21}
    \end{tabular}
    \begin{flushleft}
      \scriptsize In these figures, red points are the regression residuals of case or death
      growth rate on confounders in states without a mask mandate. Blue points are
      states with a mask mandate 14 (21 for deaths) days prior. The
      red line is the average across states without a mask mandate 14
      (21 for deaths) days earlier. The blue line is the average
      across states with a mask mandate 14 (21 for deaths) earlier.
    \end{flushleft}
  \end{minipage}
\end{figure}

### 2020-09-04 cases_and_policies/rmd/regressionsWithPlots.Rmd
updating Figure 10 with residual versions and regressionWithPlots

sgtype <- opts_knit$get("rmarkdown.pandoc.to")
sgstyle <- 'default'
colors <-  scale_color_solarized
colors_fill <- scale_fill_solarized
rootdir <- system("git rev-parse --show-toplevel", intern=TRUE)[1]
source(paste(rootdir,"cases_and_policies/R/exploratory.R", sep="/"))
source(paste(rootdir,"cases_and_policies/R/dataprep.R",sep="/"))
source(paste(rootdir,"cases_and_policies/rmd/varlabels.R", sep="/"))
source(paste(rootdir,"cases_and_policies/rmd/regprep.R", sep="/"))
source(paste(rootdir,"cases_and_policies/rmd/generatetables.R", sep="/"))
source(paste(rootdir,"cases_and_policies/rmd/bootstrap_felm.R", sep="/"))
```

@@

df$logpop <- log(df$Population.2018)
df$logsq <- log(df$Square.Miles)
df$repub <- as.factor(df$party)
df$party_dum <- ifelse(df$party == "republican", 0, 1)
df$month3 <- ifelse(df$month==3,0,1)
df$month4 <- ifelse(df$month==4,0,1)
df$month5 <- ifelse(df$month==5,0,1)
df$month6 <- ifelse(df$month==6,0,1)
L <- L.c

pols <- c("pmaskbus","pk12","pshelter","pmovie","prestaurant","pnonessential")
infovars <- list(c("dlogdc", "logdc"),
                 c("dlogdc", "logdc","dlogdc.national", "logdc.national"))
infovarsd <- list(c("dlogdd", "logdd"),
                  c("dlogdd", "logdd","dlogdd.national", "logdd.national"))
bvars <- c("workplaces","retail","grocery","transit")
vars <- c(infovars[[2]],infovarsd[[2]],pols,"pindex")
for(i in 1:length(vars)) {
  L<-L.c
  v <- sprintf("%s.L14",vars[i])
  df[,v] <- panellag(df[,vars[i]],df$state,df$date, lag=L)
  L<-L.d
  v <- sprintf("%s.L21",vars[i])
  df[,v] <- panellag(df[,vars[i]],df$state,df$date, lag=L)
}

for(i in 1:length(bvars)) {
  L<-L.c
  v <- sprintf("%s.L14",bvars[i])
  df[,v] <- panellag(df[,bvars[i]],df$state,df$date, lag=L+14)
  L<-L.d
  v <- sprintf("%s.L21",bvars[i])
  df[,v] <- panellag(df[,bvars[i]],df$state,df$date, lag=L+14)
}

svars <- c("logpop","logsq","Percent.Unemployed..2018.",
               "Percent.living.under.the.federal.poverty.line..2018.",
               "Percent.at.risk.for.serious.illness.due.to.COVID","party_dum")
mvars.c <- mvars.d <- mvars <- c("month4","month5","month6")
# mvars.c <- c("month4.c","month5.c","month6.c")
# mvars.d <- c("month4.d","month5.d","month6.d")
ijs <- expand.grid(1:length(svars), 1:length(mvars.c))
msvars.c <- vector(mode="character", length=lengths(ijs)[[1]])
msvars.d <- vector(mode="character", length=lengths(ijs)[[1]])
for (ij in (1:lengths(ijs)[[1]])){
  v <-  sprintf("%s.%s",svars[ijs[[1]][ij]],mvars.c[ijs[[2]][ij]])
  msvars.c[ij] <- v
  df[,v] <- df[,svars[ijs[[1]][ij]]]*df[,mvars.c[ijs[[2]][ij]]]
  v <-  sprintf("%s.%s",svars[ijs[[1]][ij]],mvars.d[ijs[[2]][ij]])
  msvars.d[ij] <- v
  df[,v] <- df[,svars[ijs[[1]][ij]]]*df[,mvars.d[ijs[[2]][ij]]]
}

bvars.c <- c("workplaces.L14","retail.L14","grocery.L14","transit.L14")
bvars.d <- c("workplaces.L21","retail.L21","grocery.L21","transit.L21")

wvars <- c(svars,mvars.c,msvars.c,"month3","mask_percent","logvote","dlogtests")
wvars.d <- c(svars,mvars.d,msvars.d,"month3","mask_percent","logvote")
xvars <- c(bvars.c,"dlogdc.L14","logdc.L14")
xvars.national <- c(bvars.c,"dlogdc.L14","logdc.L14","dlogdc.national.L14","logdc.national.L14")
xvars.d <- c(bvars.d,"dlogdd.L21","logdd.L21")
xvars.national.d <- c(bvars.d,"dlogdd.L21","logdd.L21","dlogdd.national.L21","logdd.national.L21")

# wvars <- c(bvars.c,svars,mvars.c,msvars.c,"month3","mask_percent","logvote","dlogtests")
# wvars.d <- c(bvars.d,svars,mvars.d,msvars.d,"month3","mask_percent","logvote")
# xvars <- c("dlogdc.L14","logdc.L14")
# xvars.national <- c("dlogdc.L14","logdc.L14","dlogdc.national.L14","logdc.national.L14")
# xvars.d <- c("dlogdd.L21","logdd.L21")
# xvars.national.d <- c("dlogdd.L21","logdd.L21","dlogdd.national.L21","logdd.national.L21")

# pols.L14 <- c("pmaskbus.L14","pk12.L14","pshelter.L14","pmovie.L14","prestaurant.L14","pnonessential.L14")
# pols.L21 <- c("pmaskbus.L21","pk12.L21","pshelter.L21","pmovie.L21","prestaurant.L21","pnonessential.L21")
pols.L14 <- c("pmaskbus.L14","pk12.L14","pshelter.L14","pindex.L14")
pols.L21 <- c("pmaskbus.L21","pk12.L21","pshelter.L21","pindex.L21")

@@

# DML2.for.PLM <- function(x, d, y, dreg, yreg, sdf, nfold=5) {
#   unit <- as.double(sdf$state)
#   j <-  quantile(unit, c(1:(nfold-1))/nfold)
#   I <- vector(mode = "list", length = nfold)
#   I[[1]] <- unit <= j[1]
#   for (b in 2:(nfold-1)) {
#     I[[b]] <- (unit <= j[b])&(unit > j[b-1])
#   }
#   I[[nfold]] <- unit > j[nfold-1]
#   nobs <- nrow(x)
#   ytil <- dtil <- rep(NA, nobs)
#   for(b in 1:nfold){
#     I1 <- (I[[b]]==FALSE)
#     I2 <- I[[b]]
#     dfit <- dreg(x[I1,], d[I1]) # take a fold out
#     yfit <- yreg(x[I1,], y[I1]) # take a fold out
#     dhat <- predict(dfit, x[I2,], type="response") #predict the fold out
#     yhat <- predict(yfit, x[I2,], type="response") #predict the fold out
#     dtil[I2] <- (d[I2] - dhat) #record residual
#     ytil[I2] <- (y[I2] - yhat) #record residual
#     cat(b," ")
#   }
# sdf$ytil <- ytil
# sdf$dtil <- dtil
# rfit <- felm(ytil ~ dtil| 0| 0 | state, data=sdf) #estimate the main parameter by regressing one residual on the other
# coef.est <- coef(rfit)[2] #extract coefficient
# se <- sqrt(vcov(rfit)[2,2])
# cat(sprintf("\ncoef (se) = %g (%g)\n", coef.est , se))
# return( list(coef.est = coef.est , se=se) )
# }

@@

p <- "pmaskbus"
figdev(sprintf("%s/tex/tables_and_figures/%s-cases-%s.pdf", rootdir, p, L.c), width=4, height=3)
print(fig)
dev.off()
fig
```

>>>>

sgtype <- opts_knit$get("rmarkdown.pandoc.to")
sgstyle <- 'default'
colors <-  scale_color_solarized
colors_fill <- scale_fill_solarized
rootdir <- system("git rev-parse --show-toplevel", intern=TRUE)[1]

source(paste(rootdir,"cases_and_policies/R/exploratory.R", sep="/"))
source(paste(rootdir,"cases_and_policies/R/dataprep.R",sep="/"))
source(paste(rootdir,"cases_and_policies/rmd/varlabels.R", sep="/"))
source(paste(rootdir,"cases_and_policies/rmd/regprep.R", sep="/"))
source(paste(rootdir,"cases_and_policies/rmd/generatetables.R", sep="/"))
source(paste(rootdir,"cases_and_policies/rmd/bootstrap_felm.R", sep="/"))
```

@@

## the following part is moved to regprep.R
# df$logpop <- log(df$Population.2018)
# df$logsq <- log(df$Square.Miles)
# df$repub <- as.factor(df$party)
# df$party_dum <- ifelse(df$party == "republican", 0, 1)
# df$month3 <- ifelse(df$month==3,0,1)
# df$month4 <- ifelse(df$month==4,0,1)
# df$month5 <- ifelse(df$month==5,0,1)
# df$month6 <- ifelse(df$month==6,0,1)
# L <- L.c 
# pols <- c("pmaskbus","pk12","pshelter","pmovie","prestaurant","pnonessential")
# infovars <- list(c("dlogdc", "logdc"),
#                  c("dlogdc", "logdc","dlogdc.national", "logdc.national"))
# infovarsd <- list(c("dlogdd", "logdd"),
#                   c("dlogdd", "logdd","dlogdd.national", "logdd.national"))
# bvars <- c("workplaces","retail","grocery","transit")
# vars <- c(infovars[[2]],infovarsd[[2]],pols,"pindex")
# for(i in 1:length(vars)) {
#   L<-L.c
#   v <- sprintf("%s.L14",vars[i])
#   df[,v] <- panellag(df[,vars[i]],df$state,df$date, lag=L)
#   L<-L.d
#   v <- sprintf("%s.L21",vars[i])
#   df[,v] <- panellag(df[,vars[i]],df$state,df$date, lag=L)
# } 
# for(i in 1:length(bvars)) {
#   L<-L.c
#   v <- sprintf("%s.L14",bvars[i])
#   df[,v] <- panellag(df[,bvars[i]],df$state,df$date, lag=L+14)
#   L<-L.d
#   v <- sprintf("%s.L21",bvars[i])
#   df[,v] <- panellag(df[,bvars[i]],df$state,df$date, lag=L+14)
# } 
# svars <- c("logpop","logsq","Percent.Unemployed..2018.",
#                "Percent.living.under.the.federal.poverty.line..2018.",
#                "Percent.at.risk.for.serious.illness.due.to.COVID","party_dum")
# mvars.c <- mvars.d <- mvars <- c("month4","month5","month6")
# # mvars.c <- c("month4.c","month5.c","month6.c")
# # mvars.d <- c("month4.d","month5.d","month6.d")
# ijs <- expand.grid(1:length(svars), 1:length(mvars.c))
# msvars.c <- vector(mode="character", length=lengths(ijs)[[1]])
# msvars.d <- vector(mode="character", length=lengths(ijs)[[1]])
# for (ij in (1:lengths(ijs)[[1]])){
#   v <-  sprintf("%s.%s",svars[ijs[[1]][ij]],mvars.c[ijs[[2]][ij]])
#   msvars.c[ij] <- v
#   df[,v] <- df[,svars[ijs[[1]][ij]]]*df[,mvars.c[ijs[[2]][ij]]]
#   v <-  sprintf("%s.%s",svars[ijs[[1]][ij]],mvars.d[ijs[[2]][ij]])
#   msvars.d[ij] <- v
#   df[,v] <- df[,svars[ijs[[1]][ij]]]*df[,mvars.d[ijs[[2]][ij]]]
# } 
# bvars.c <- c("workplaces.L14","retail.L14","grocery.L14","transit.L14")
# bvars.d <- c("workplaces.L21","retail.L21","grocery.L21","transit.L21")  
# wvars <- c(svars,mvars.c,msvars.c,"month3","mask_percent","logvote","dlogtests")
# wvars.d <- c(svars,mvars.d,msvars.d,"month3","mask_percent","logvote")
# xvars <- c(bvars.c,"dlogdc.L14","logdc.L14")
# xvars.national <- c(bvars.c,"dlogdc.L14","logdc.L14","dlogdc.national.L14","logdc.national.L14")
# xvars.d <- c(bvars.d,"dlogdd.L21","logdd.L21")
# xvars.national.d <- c(bvars.d,"dlogdd.L21","logdd.L21","dlogdd.national.L21","logdd.national.L21") 
# pols.L14 <- c("pmaskbus.L14","pk12.L14","pshelter.L14","pindex.L14")
# pols.L21 <- c("pmaskbus.L21","pk12.L21","pshelter.L21","pindex.L21")

@@

@@

p <- "pmaskbus"
figdev(sprintf("%s/tex/tables_and_figures/%s-cases-%s.pdf", rootdir, p, L.c), width=4, height=3)
print(fig)
dev.off()
fig

# use residuals after controlling for confounders 
wv <- c(svars,mvars.c,msvars.c,"month3") 
fmla <- as.formula(paste("dlogdc", "~",  paste(wv, collapse=" + "), sep=" "))
g1 <- predict(lm(fmla, data=df))
df$dlogdc.res <- df[,yv] - g1    
fig <- pfigure(df, "pmaskbus", stlab=-1,outcome="dlogdc.res",L=L.c) + ggtitle(TeX(relabel("dlogdc given pmaskbus"))) + ylab(TeX(relabel("residual of dlogdc"))) + 
  ylim(c(-1.8,0.5))  
p <- "pmaskbus"
figdev(sprintf("%s/tex/tables_and_figures/%s-cases-res-%s.pdf", rootdir, p, L.c), width=4, height=3)
print(fig)
dev.off()
fig

wv <- c(svars,mvars.d,msvars.d,"month3") 
fmla <- as.formula(paste("dlogdd", "~",  paste(wv, collapse=" + "), sep=" "))
g1 <- predict(lm(fmla, data=df))
df$dlogdd.res <- df[,yv] - g1  
fig <- pfigure(df, "pmaskbus", stlab=-1,outcome="dlogdd.res",L=L.d) + ggtitle(TeX(relabel("dlogdd given pmaskbus"))) + ylab(TeX(relabel("residual of dlogdd")))+ 
  ylim(c(-0.9,0.5))  
p <- "pmaskbus"
figdev(sprintf("%s/tex/tables_and_figures/%s-deaths-res-%s.pdf", rootdir, p, L.d), width=4, height=3)
print(fig)
dev.off()
fig

### 2020-09-04 tex/covidEffectOfNPI-revision.tex
updating Figure 10 with residual versions and regressionWithPlots

\begin{figure}[ht]
  \caption{Case and death growth conditional on mask mandates \label{fig:masks}}\vspace{0.2cm}
 \begin{minipage}{\linewidth}
    \centering
    \begin{tabular}{cc}
      \includegraphics[width=0.5\textwidth]{tables_and_figures/pmaskbus-cases-14}
      &
        \includegraphics[width=0.5\textwidth]{tables_and_figures/pmaskbus-deaths-21}
    \end{tabular}
    \begin{flushleft}
      \scriptsize In these figures, red points are the case or death
      growth rate in states without a mask mandate. Blue points are
      states with a mask mandate 14 (21 for deaths) days prior. The
      red line is the average across states without a mask mandate 14
      (21 for deaths) days earlier. The blue line is the average
      across states with a mask mandate 14 (21 for deaths) earlier.
    \end{flushleft}
  \end{minipage}
\end{figure}

@@

We now analyze how behavior and policies together influence case and death
growth rates. We begin with some simple graphical evidence of the
effect of policies on case and death growth. Figure \ref{fig:masks}
shows average case and death growth conditional on date and whether
masks are mandatory for employees.\footnote{We take 14 and 21 day lags of mask policies for case and death growths, respectively, to  identify the states with a mask mandate because policies affect cases and deaths with time lags. See our discussion in the Appendix \ref{sec:timing}.  } The left panel of the figure shows
that states with a mask mandate consistently have 0-0.2 lower case
growth than states without. The right panel  also illustrates that
states with a mask mandate tend to have lower average death growth than states without a mask mandate.

>>>>

\afterpage{
 \clearpage
\begin{figure}[ht]
  \caption{Case and death growth conditional on mask mandates \label{fig:masks}}\vspace{0.2cm}
 \begin{minipage}{\linewidth}
    \centering
    \begin{tabular}{cc}
      \includegraphics[width=0.5\textwidth]{tables_and_figures/pmaskbus-cases-14}
      &
        \includegraphics[width=0.5\textwidth]{tables_and_figures/pmaskbus-deaths-21}
    \end{tabular}
    \begin{flushleft}
      \scriptsize In these figures, red points are the case or death
      growth rate in states without a mask mandate. Blue points are
      states with a mask mandate 14 (21 for deaths) days prior. The
      red line is the average across states without a mask mandate 14
      (21 for deaths) days earlier. The blue line is the average
      across states with a mask mandate 14 (21 for deaths) earlier.
    \end{flushleft}
  \end{minipage}\\ \medskip
 \begin{minipage}{\linewidth}
    \centering
    \begin{tabular}{cc}
      \includegraphics[width=0.5\textwidth]{tables_and_figures/pmaskbus-cases-res-14}
      &
        \includegraphics[width=0.5\textwidth]{tables_and_figures/pmaskbus-deaths-res-21}
    \end{tabular}
    \begin{flushleft}
      \scriptsize In these figures, red points are the regression residuals of case or death
      growth rate on confounders in states without a mask mandate. Blue points are
      states with a mask mandate 14 (21 for deaths) days prior. The
      red line is the average across states without a mask mandate 14
      (21 for deaths) days earlier. The blue line is the average
      across states with a mask mandate 14 (21 for deaths) earlier.
    \end{flushleft}
  \end{minipage}
\end{figure} 
\clearpage}

@@

We now analyze how behavior and policies together influence case and death
growth rates. We begin with some simple graphical evidence of the
effect of policies on case and death growth. The upper two panels of figure \ref{fig:masks}
show average case and death growth conditional on date and whether
masks are mandatory for employees.\footnote{We take 14 and 21 day lags of mask policies for case and death growths, respectively, to  identify the states with a mask mandate because policies affect cases and deaths with time lags. See our discussion in the Appendix \ref{sec:timing}.} The upper left panel of the figure shows that states with a mask mandate consistently have 0-0.2 lower case
growth than states without. The upper right panel also illustrates that
states with a mask mandate tend to have lower average death growth than states without a mask mandate. The lower two panels of figure \ref{fig:masks} are similar to the upper two panels except that we use the residuals of  regressing case or death growth on state-level characteristics, month dummies, and their interactions in place of  case or death growth, indicating that states with a mask mandate tend to have lower case and death growths even after controlling for confounders.
